Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-02-18 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: small changes

Further expand the Kerr expression as:
%
\begin{subequations}
\begin{align}
\omega_{\iota} &= (1/M) \sum_{j = 0}^{N} \, \chi^{j} \omega^{(j)}_{k} \, \left( 1 + \gamma \delta \omega^{(j)}_{k} \right), \\
\tau_{\iota}   &= M     \sum_{j = 0}^{N} \, \chi^{j} \tau^{(j)}_{k}   \, \left( 1 + \gamma \delta \tau^{(j)}_{k} \right),
\end{align}
\label{eq:kerr_expansion}
\end{subequations}

Also,
%
\begin{equation}
\gamma
%
= \frac{\ell^{p}}{M_{\rm s}^{p}}
%
% = \frac{\ell^{p} (1 + z)^{p}}{M^{p}}
%
= \left[
\frac{\ell c^2 (1 + z)}{G M}
\right]^{p}
\label{eq:def_gamma}
\end{equation}

@@

In this formalism our QNM frequencies are expressed through functions,
%
\begin{subequations}
\begin{align}
\omega_k &= \omega_k(m_1, m_2, \chi_1, \chi_2,\ell, \{\delta \omega_k^{(j)}\})\\
\tau_k   &= \tau _k(m_1, m_2, \chi_1, \chi_2, \ell, \{\delta \tau_k^{(j)}\})
\end{align}
\end{subequations}
%
where the estimates of $(m_1,\, m_2,\, \chi_1,\, \chi_1)$ are used to predict the final mass and spin $(M_f,\, \chi_f)$ through~\cite{Taracchini:2013rva,Hofmann:2016yih}.
%
Additionally, $(\omega_k,\, \tau_k)$ depend on the coupling constant $\ell$ defined in Eq.~\eqref{eq:def_gamma} as well as the sets, $\{\delta \omega_k^{(j)}\},$ and $\{\delta \tau_k^{(j)}\}$, where $j$ ranges from 0 to the power of spin in Eqs.~\ref{eq:kerr_expansion} upto which we include corrections.
%
For example, if we restrict ourselves to corrections in the $j=0$, or $\chi_f^0$, terms for just the $k=(2,2,0)$ mode, then we would just have two additional degrees of freedom, $(\delta \omega_0^{(0)}, \delta \tau_0^{(0)})$. In this paper, we would restrict ourselves to corrections to the $j=0$ and $j=1$ terms, i.e., leading and next-to-leading order corrections to the spin, $\chi_f$.
%

@@

Just as in GR where the coefficients $\omega^{j}$, $\tau^{j}$ are determined by
comparison against the Kerr QNMs, a theory-specific calculation predicts the
numerical values of $\delta\omega^{(j)}$ and $\delta\tau^{(j)}$
%
This means that using the $\pSEOB$ waveform model with theory-specific informed
QNMs we reduce the parameters space of beyond-GR parameters to one, i.e., to
only $\gamma$, which in turn contains information about the length scale $\ell$.
%
It is important to emphasize that our procedure is different from that of~\cite{Carullo:2021dui}
which, for a given value of $p$, varied all $\ell$, $\delta\omega^{(i)}$ and $\delta\tau^{(i)}$ parameters.
%
% As we will see later, our theory-informed approach can result to stronger constraints to certain modified ....

Now as we have seen in Sec.~\ref{sec:review_theories}, the QNM of slowly-rotating BHs
is modified gravity theories come in two families depending on their transformation upon a parity
transformation: axial and polar. Which is the one we use to match against Eqs.~\eqref{eq:qnm_nongr_iso}?
%
To answer this question one has to work on a theory-by-theory case and in each theory perform a
a translation between the metric perturbations $h_{\mu\nu}$ in the Regger-Wheeler-Zerilli gauge and
connect it with the transverse-traceless gauge used to described GWs (see e.g.~Ref.~\cite{Maggiore:2018sht}, Sec. xxx),
%
In GR, both axial and polar QNMs are the same (a property known as \emph{isospectrality}) and therefore which
QNM we use to model the ringdown makes no difference.
%
In beyond-GR theories, isospectrality is in general broken (see the discussion in Ref.~\cite{Hui:2021cpm} for a counterexample).
and thus how axial and polar gravitational QNMs appear in the GW signal has to be answered in theory-by-theory.
%
This is outside the scope of this paper and here we take the more pragmatic approach
of simply choosing the \emph{least damped gravitational mode between the two parities.}
%
The justification is this is the mode which, if excited during the merger, is
the most likely to appear in the signal.

@@

Assuming Gaussian and stationary noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln P(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk denotes
complex conjugation and $S_{n}(f)$ is the power spectrum density of the
detector.
%
Assuming a specific prior distribution for our parameters, we stochastically
sample over the parameter space using a \texttt{LALInference} Markov Chain
Monte Carlo algorithm.~\cite{lalsuite}. We subsequently marginalize over the
remaining parameters to obtain the posterior probability distribution on
$\ell$. An example of this is shown in Fig. XX.

>>>>

Further expand the Kerr expression as:
%
\begin{subequations}
\begin{align}
\omega_{J} &= (1/M) \sum_{j = 0}^{N} \, \chi^{j} \omega^{(j)}_{k} \, \left( 1 + \gamma \delta \omega^{(j)}_{k} \right), \\
\tau_{J}   &= M     \sum_{j = 0}^{N} \, \chi^{j} \tau^{(j)}_{k}   \, \left( 1 + \gamma \delta \tau^{(j)}_{k} \right),
\end{align}
\label{eq:kerr_expansion}
\end{subequations}

Also,
%
\begin{equation}
\gamma
%
= \frac{\ell^{p}}{M_{\rm s}^{p}}
%
= \left[
\frac{\ell c^2 (1 + z)}{G M}
\right]^{p}
\label{eq:def_gamma}
\end{equation}

@@

In this formalism our QNM frequencies are expressed through functions,
%
\begin{subequations}
\begin{align}
\omega_k &= \omega_k(m_1, m_2, \chi_1, \chi_2,\ell, \{\delta \omega_k^{(j)}\})\\
\tau_k   &= \tau _k(m_1, m_2, \chi_1, \chi_2, \ell, \{\delta \tau_k^{(j)}\})
\end{align}
\end{subequations}
%
where the estimates of $(m_1,\, m_2,\, \chi_1,\, \chi_1)$ are used to predict the final mass and spin $(M_f,\, \chi_f)$ through~\cite{Taracchini:2013rva,Hofmann:2016yih}.
%
Additionally, $(\omega_k,\, \tau_k)$ depend on the coupling constant $\ell$
defined in Eq.~\eqref{eq:def_gamma} as well as the sets, $\{\delta \omega_k^{(j)}\},$ and
$\{\delta \tau_k^{(j)}\}$, where $j$ ranges from zero to the power of spin in Eqs.~\eqref{eq:kerr_expansion}
up to which we include corrections.
%
For example, if we restrict ourselves to corrections in the $j=0$, or $\chi_f^0$, terms for just the $k=(2,2,0)$ mode, then we would just have two additional degrees of freedom, $(\delta \omega_0^{(0)}, \delta \tau_0^{(0)})$. In this paper, we would restrict ourselves to corrections to the $j=0$ and $j=1$ terms, i.e., leading and next-to-leading order corrections to the spin, $\chi_f$.
%

@@

Just as in GR where the coefficients $\omega^{j}$, $\tau^{j}$ are determined by
comparison against the Kerr QNMs, a theory-specific calculation predicts the
numerical values of $\delta\omega^{(j)}$ and $\delta\tau^{(j)}$
%
This means that using the $\pSEOB$ waveform model with theory-specific informed
QNMs we reduce the parameters space of beyond-GR parameters to one, i.e., to
only $\gamma$, which in turn contains information about the length scale $\ell$.
%
It is important to emphasize that our procedure is different from that of Ref.~\cite{Carullo:2021dui}
which, for a given value of $p$, varied all $\ell$, $\delta\omega^{(i)}$ and $\delta\tau^{(i)}$ parameters.
%
% As we will see later, our theory-informed approach can result to stronger constraints to certain modified ....

As we have seen in Sec.~\ref{sec:review_theories}, the QNM of slowly-rotating BHs
is modified gravity theories come in two families depending on their transformation upon a parity
transformation: axial and polar. Which is the one we use to match against Eqs.~\eqref{eq:qnm_nongr_iso}?
%
To answer this question one has to work on a theory-by-theory case and in each theory perform a
a translation between the metric perturbations $h_{\mu\nu}$ in the Regger-Wheeler-Zerilli gauge and
connect it with the transverse-traceless gauge used to described GWs (see e.g.~Ref.~\cite{Maggiore:2018sht}, Sec. xxx),
%
In GR, both axial and polar QNMs are the same (a property known as \emph{isospectrality}) and therefore which
QNM we use to model the ringdown makes no difference.
%
In beyond-GR theories, isospectrality is in general broken (see the discussion in Ref.~\cite{Hui:2021cpm} for a counterexample).
and thus how axial and polar gravitational QNMs appear in the GW signal has to be answered in theory-by-theory.
%
This is outside the scope of this paper and here we take the more pragmatic approach
of simply choosing the \emph{least damped gravitational mode between the two parities.}
%
The justification is this is the mode which, if excited during the merger, is
the most likely to appear in the signal.

@@

Assuming Gaussian and stationary noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln P(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk denotes
complex conjugation and $S_{n}(f)$ is the power spectrum density of the
detector.
%
Assuming a specific prior distribution for our parameters, we stochastically
sample over the parameter space using a \texttt{LALInference} Markov-Chain
Monte Carlo algorithm.~\cite{lalsuite}. We subsequently marginalize over the
remaining parameters to obtain the posterior probability distribution
function~(PDF) on $\ell$. An example of this is shown in Fig. XX.
